---
title: "MSBA Math Primer: Module 1"
author: "Burcu Eke Rubini^[University of New Hampshire, Burcu.EkeRubini@unh.edu]"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Learn how to solve systems of linear equations in R.
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(Lahman)
library(datasets)
library(fBasics)
library(base)
tutorial_options(exercise.timelimit = 60)
knitr::opts_chunk$set(error = TRUE)
```

## Linear Algrebra

This tutorial is designed to review basic operations on vectors and matrices. Throughout this tutorial lower case bold letters, such as **a** denotes vectors, upper case bold letters, such as **A** denotes matrices, and lower case letters, such as x denotes scalars.

For example, a $3 \times 1$ vector **b** can be denoted as:

$$b=\begin{pmatrix}
b_1\\
b_2\\
b_3 \end{pmatrix}$$

###

Or, a $3 \times 2$ vector **A** can be denoted as:

$$A=\begin{bmatrix}
a_{11} & a_{12}\\
a_{21} & a_{22}\\
a_{31} & a_{32} \end{bmatrix}$$

where $a_{ij}$ is the element on $i$-th row and $j$-th column

## Basic Terminology

### Basic Definitions

- If a matrix has the same number of rows and number of columns, that matrix is classified as *square* matrix.
- A *diagonal* matrix is a matrix where all off diagonal values are 0. (Diagonal refers to $a_{ii}$)
- If a square diagonal matrix has 1's in its diagonals, it is called an *identity* matrix, and denoted as $I_n$, where $n$ is the number of columns (=number of rows)

$$I_3=\begin{bmatrix}
1&0&0\\
0&1&0\\
0&0&1\end{bmatrix}$$

- Transpose of a matrix is a matrix when the rows and columns are switched.

$$A=\begin{bmatrix}
a_{11} & a_{12}\\
a_{21} & a_{22}\\
a_{31} & a_{32} \end{bmatrix} \text{ and } A^T=A'=\begin{bmatrix}
a_{11} & a_{21} &a_{31}\\
a_{12} & a_{22} & a_{32} \end{bmatrix}$$

## Addition

### Adding two or more matrices

As long as they are of the same dimension, we can do element-by-element addition using matrices and vectors. For example:

$$A+B=\begin{bmatrix}
1 & -1\\
2 & 2\\
3 & -3 \end{bmatrix} + \begin{bmatrix}
1 &1\\
1 &1\\
1 &1 \end{bmatrix} = \begin{bmatrix}
2 & 0\\
3 & 3\\
4 & -2 \end{bmatrix} $$


### Example


The R code required to add the two matrices seen above (A and B are predefined):

```{r echo=FALSE}
A=matrix(c(1:3, -1,2,-3), nrow=3, ncol = 2)
B=matrix(rep(1, 6), 3, 2)
C=matrix(c(rep(1, 4), -1,-2), 2, 3)
D=matrix(c(1:3, -1,2,-3, 1, 1, 1), nrow=3, ncol = 3)
I=diag(3)
E=matrix(1:4, nrow=2, ncol=2)
F=matrix(2:5, nrow=2, ncol=2)
```

```{r A-plus-B, exercise=TRUE}
A+B
```


### Exercise 2

Now calculate B+A. Do you get the same results as in the example?

```{r set1ex2, exercise=TRUE, exercise.lines = 5}

```

```{r set1ex2-solution}
A+B
```

### Exercise 3

Now write a code that adds the matrices A and C. Can you do it? If not, what is the problem? Try printing A and C.

```{r set1ex3, exercise=TRUE, exercise.lines = 5}

```

```{r set1ex3-solution}
A+C
```

```{r set1ex3-check}
"We cannot add matrices of different dimensions"
```

### Exercise 4

How about $$A^T+C$$? Can we add those two?

```{r set1ex4, exercise=TRUE, exercise.lines = 5}

```

<div id="set1ex4-hint">
**Hint:** You can take transpose of matrix B by `t(B)`.
</div>

```{r set1ex4-solution}
t(A)+C
```

```{r set1ex4-check}
"Yes! Now they both have the same number of rows and columns"
```


### Properties of Addition

Assuming **A**, **B** and **C** are matrices with the same dimensions:

- $(A+B) +C=A+ (B+C)$
- $A+B=B+A$
- $A+O=O+A=A$
- $A+ (-A) = (-A) +A=O$

## Multiplication

### Multiplying two matrices

Two  matrices **A** and **B** can  be  multiplied  together  if  the  number  of  rows  in **B** equals  the  number  of columns in **A**: $A_{m \times n}B_{n \times u} = C_{m \times u}$. An element $c_{ik}$ is calculated as  $c_{ik}=\sum_{i=1}^n a_{ij}b_{jk}$


### Multiplication Example 
To illustrate:

$$\begin{eqnarray}
AC &=&\begin{bmatrix}
1 & -1\\
2 & 2\\
3 & -3 \end{bmatrix}  \begin{bmatrix}
1 &1&-1\\
1 &1&-2\\ \end{bmatrix} \\
&=& \begin{bmatrix}
(1)(1)+(-1)(1) & (1)(1)+(1)(-1) & (1)(-1) +(-1)(-2) \\
(2)(1) +(2)(1) & (2)(1) +(2)(1) & (2)(-1) +(2)(-2)\\
(3)(1) + (-3)(1) & (3)(1) + (-3)(1) & (3)(-1) + (-3)(-2) \end{bmatrix} \\
&=&  \begin{bmatrix}
0 & 0 & 1\\
4 & 4 & -6 \\
0 & 0 & 3
\end{bmatrix}
\end{eqnarray}$$

Let's check our results in R. `*` is reserved for scalar multiplication. For Matrix multiplication, we can use `%*%`:

```{r A-times-C, exercise=TRUE}
A %*% C
```


### Exercise 5

Multiply the matrix **D** by matrix **A**, $DA$ where $$D=\begin{bmatrix}
1 & -1 & 1\\
2 & 2 & 1\\
3 & -3 & 1 \end{bmatrix}$$

Now, check your work in R. Can you get the same results for $AD$ and $DA$?

```{r set1ex5, exercise=TRUE}

```

```{r set1ex5-solution}
D%*%A
```

### Exercise 6

Multiply the matrix **D** by matrix **I**, $DI$ where $I=I_3$. Do you get the same results for $ID$ and $DI$?

```{r set1ex6, exercise=TRUE}

```

```{r set1ex6-solution}
D %*% I
```
    
### Exercise 7

Now, multiply the matrix **I** by matrix **D**, $ID$ where $I=I_3$. Do you get the same results for $ID$ and $DI$?

```{r set1ex7, exercise=TRUE}

```

```{r set1ex7-solution}
I %*% D
```


```{r set1ex7-check}
"Good Job! ID=DI!"
```

###

For any $n \times m$ matrix **A** multiplied by the identity matrix is equal to the matrix itself.

- $AI_m=A$
- $I_nA=A$
- When $m=n$, $AI_n=I_nA=A$

### Exercise 8

Consider the two matrices E and F (matrices already in R). Calculate $(EF)^T$.


```{r set1ex8, exercise=TRUE, exercise.lines = 5}
print(E)
print(F)

```

```{r set1ex8-solution}
print(E)
print(F)
t(E%*%F)
```

### Exercise 9

Consider the two matrices E and F. Calculate $F^TE^T$.


```{r set1ex9, exercise=TRUE, exercise.lines = 5}

```

```{r set1ex9-solution}
t(F)%*%t(E)
```

```{r set1ex9-check}
"Notice that exerices 8 and 9 yield the same output!"
```

###

Identity: $$ (AB)^T = B^T A^T$$


## Quiz 1

### Quiz 1 Questions

```{r quiz 1}
quiz(
  question("Which one of the following identities / operations require the matrices to be of the same dimension $(n \times m$)?",
    answer("(AB)^T = B^T A^T"),
    answer("(AB)^T = AB"),
    answer("A+B", correct = TRUE),
    answer("AB")
  )
)
```

